Dependent Linear Type Theory for Resource Aware Certified Programming

### 2021-09-08 writing/draft.tex
M10

% packages
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{syntax}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{listings}

\newcommand{\LNLD}{\text{LNL}_\text{D}}

@@

  We present a full spectrum dependent linear type theory that provides a more direct approach to enforce linearity, eschewing the use of modality operators or annotations. We believe that linearity is an intrinsic property completely determined by the type of a term. To accomplish this, our type system uses two class of type universes in the style of Krishnaswami, $L$ for linear and $U$ for non-linear. Due to the abandonment of modalities, we use two distinct functions types $(x : A) \rightarrow B$ for non-linear functions and $A \multimap B$ for linear functions in the style of Pfenning. We formalize an operational semantics and prove the type soundness of our system. To further demonstrate the applications of dependent linear types, we extend our type system with imperative programming features, leveraging both dependency and linearity for internalized certified imperative programming. We augment our soundness theorem to account for these extensions, showing that imperative resources are properly managed. We have implemented a prototype of our language extended with other features for practicality and ergonomics.

@@

>>>>

% packages
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{syntax}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{float}

\floatstyle{boxed} 
\restylefloat{figure}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{10pt}
\setlength{\grammarindent}{16em}

\newcommand{\LNLD}{\text{LNL}_\text{D}}
\newcommand{\ok}{\textbf{ok}}
\newcommand{\pure}{\textbf{pure}}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\U}{\textbf{U}}
\renewcommand{\L}{\textbf{L}}
\newcommand{\pstep}{\leadsto_p}

@@

  We present a full spectrum dependent linear type theory that provides a more direct approach to enforce linearity, eschewing the use of modality operators or annotations. We believe that linearity is an intrinsic property completely determined by the type of a term. To accomplish this, our type system uses two class of type universes in the style of Krishnaswami, $L$ for linear and $U$ for non-linear. Due to the abandonment of modalities, we use two distinct functions types $(x : A) \rightarrow B$ for non-linear functions and $(x : A) \multimap B$ for linear functions in the style of Pfenning. We formalize an operational semantics and prove the type soundness of our system. To further demonstrate the applications of dependent linear types, we extend our type system with imperative programming features, leveraging both dependency and linearity for internalized certified imperative programming. We augment our soundness theorem to account for these extensions, showing that imperative resources are properly managed. We have implemented a prototype of our language extended with other features for practicality and ergonomics.

@@

  \begin{figure}[h]
    \caption{Syntax}
    \begin{grammar}
      <$x, y, z, f, g, h$> $\in$ variables

      <$m, n, A, B, C$> $\in$ expressions

      <$v$> $\in$ values

      <expressions> ::= $\U$ | $\L$ | $x$
      \alt $(x : A) \rightarrow B$ | $A \rightarrow B$
      \alt $(x : A) \multimap B$ | $A \multimap B$
      \alt $\lambda x. n$ | $m\ n$

      <values> ::= $\U$ | $\L$ | $x$
      \alt $(x : A) \rightarrow B$ | $A \rightarrow B$
      \alt $(x : A) \multimap B$ | $A \multimap B$
      \alt $\lambda x. n$
    \end{grammar}
  \end{figure}

  \begin{figure}[h]
    \caption{Structural Judgments}
    \begin{mathpar}
      \inferrule
      { }
      { \cdot\ ; \cdot \ok }

      \inferrule
      { \Gamma ; \Delta\ \ok \\ \Gamma ; \Delta \vdash A : \U }
      { \Gamma, x : A ; \Delta\ \ok }

      \inferrule
      { \Gamma ; \Delta\ \ok \\ \Gamma ; \Delta \vdash A : \L }
      { \Gamma ; \Delta, x : A\ \ok }

    \end{mathpar}
  \end{figure}

  \begin{figure}[h]
    \caption{Type Formation} 
    \begin{mathpar}
      \inferrule
      { }
      { \Gamma ; \cdot \vdash \U : \U }

      \inferrule
      { }
      { \Gamma ; \cdot \vdash \L : \U }

      \inferrule
      { \Gamma ; \cdot \vdash A : \U }
      { \Gamma ; \cdot \vdash A\ \type }

      \inferrule
      { \Gamma ; \cdot \vdash A : \L }
      { \Gamma ; \cdot \vdash A\ \type }
      \\

      \inferrule
      { \Gamma ; \cdot \vdash A : \U \\ \Gamma, x : A ; \cdot \vdash B\ \type }
      { \Gamma ; \cdot \vdash (x : A) \rightarrow B : \U }

      \inferrule
      { \Gamma ; \cdot \vdash A : \L \\ \Gamma ; \cdot \vdash B\ \type }
      { \Gamma ; \cdot \vdash A \rightarrow B : \U } 
      \\

      \inferrule
      { \Gamma ; \cdot \vdash A : \U \\ \Gamma, x : A ; \cdot \vdash B\ \type }
      { \Gamma ; \cdot \vdash (x : A) \multimap B : \L }

      \inferrule
      { \Gamma ; \cdot \vdash A : \L \\ \Gamma ; \cdot \vdash B\ \type }
      { \Gamma ; \cdot \vdash A \multimap B : \L } 
    \end{mathpar}
  \end{figure}

  \begin{figure}[h]
    \caption{Term Formation}
    \begin{mathpar}
      \inferrule
      { }
      { \Gamma, x : A ; \cdot \vdash x : A }

      \inferrule
      { }
      { \Gamma ; x : A \vdash x : A }

      \inferrule
      { \Gamma ; \Delta \vdash m : A \\ \Gamma ; \Delta \vdash A \equiv B }
      { \Gamma ; \Delta \vdash m : B } 
      \\

      \inferrule
      { \Gamma ; \cdot \vdash (x : A) \rightarrow B : \U \\ 
        \Gamma, x : A ; \cdot \vdash n : B }
      { \Gamma ; \cdot \vdash \lambda x . n : (x : A) \rightarrow B }

      \inferrule
      { \Gamma ; \cdot \vdash A \rightarrow B : \U \\
        \Gamma ; x : A \vdash n : B }
      { \Gamma ; \cdot \vdash \lambda x . n : A \rightarrow B }
      \\

      \inferrule
      { \Gamma ; \cdot \vdash (x : A) \multimap B : \L \\ 
        \Gamma, x : A ; \Delta \vdash n : B }
      { \Gamma ; \Delta \vdash \lambda x . n : (x : A) \multimap B }

      \inferrule
      { \Gamma ; \cdot \vdash A \multimap B : \L \\ 
        \Gamma ; \Delta, x : A \vdash n : B }
      { \Gamma ; \Delta \vdash \lambda x . n : A \multimap B }
      \\

      \inferrule
      { \Gamma ; \Delta_1 \vdash m : (x : A) \rightarrow B \\
        \Gamma ; \Delta_2 \vdash n : A }
      { \Gamma; \Delta_1, \Delta_2 : m\ n : [n/x]B }

      \inferrule
      { \Gamma ; \Delta_1 \vdash m : A \rightarrow B \\
        \Gamma ; \Delta_2 \vdash n : A }
      { \Gamma; \Delta_1, \Delta_2 : m\ n : B }
      \\

      \inferrule
      { \Gamma ; \Delta_1 \vdash m : (x : A) \multimap B \\
        \Gamma ; \Delta_2 \vdash n : A }
      { \Gamma; \Delta_1, \Delta_2 : m\ n : [n/x]B }

      \inferrule
      { \Gamma ; \Delta_1 \vdash m : A \multimap B \\
        \Gamma ; \Delta_2 \vdash n : A }
      { \Gamma; \Delta_1, \Delta_2 : m\ n : B }
    \end{mathpar}
  \end{figure}

  \begin{figure}[h]
    \caption{Equality}
    \begin{mathpar}
      \inferrule
      { \Gamma ; \Delta \vdash m_1 : A \\ \Gamma ; \Delta \vdash m_2 : A \\ 
        m_1 \pstep^* n \\ m_2 \pstep^* n }
      { \Gamma ; \Delta \vdash m_1 \equiv m_2 }
    \end{mathpar}
  \end{figure}

  \begin{figure}[h]
    \caption{Parallel Reduction}
    \begin{mathpar}
      \inferrule
      { }
      { n \pstep n }

      \inferrule
      { n \pstep n' }
      { \lambda x.n \pstep \lambda x.n' }

      \inferrule
      { m \pstep m' \\ n \pstep n' }
      { m\ n \pstep m'\ n' }

      \inferrule
      { n \pstep n' \\ v \pstep v' }
      { (\lambda x.n)\ v \pstep [v'/x]n' }
      \\

      \inferrule
      { A \pstep A' \\ B \pstep B' }
      { (x : A) \rightarrow B \pstep (x : A') \rightarrow B' }

      \inferrule
      { A \pstep A' \\ B \pstep B' }
      { A \rightarrow B \pstep A' \rightarrow B' }
      \\

      \inferrule
      { A \pstep A' \\ B \pstep B' }
      { (x : A) \multimap B \pstep (x : A') \multimap B' }

      \inferrule
      { A \pstep A' \\ B \pstep B' }
      { A \multimap B \pstep A' \multimap B' }
    \end{mathpar}
  \end{figure}
